---
title: "R Implementation — Waymo Unsupervised Anomaly Detection"
output: html_document
date: "2026-03-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Code Implementation

```{r}
library(tidyverse)
library(caret)
library(isotree)
library(e1071)
library(dbscan)
library(gridExtra)
library(car)
library(showtext)
set.seed(42)

# --- Waymo Font (Latin Modern Roman — the actual LaTeX font) ---
font_add("LM Roman",
         regular  = "~/Library/Fonts/lmroman10-regular.otf",
         bold     = "~/Library/Fonts/lmroman10-bold.otf",
         italic   = "~/Library/Fonts/lmroman10-italic.otf")
showtext_auto()
showtext_opts(dpi = 300)

# --- Waymo Color Palette ---
waymo_green <- "#00C389"
waymo_blue  <- "#1A73E8"
waymo_grey  <- "#8A8F94"
waymo_navy  <- "#1F233F"
waymo_white <- "#FFFFFF"

waymo_2col    <- c("With Stationary"    = waymo_blue,  "Without Stationary" = waymo_green)
waymo_3col    <- c("Cyclist"            = waymo_grey,  "Pedestrian"         = waymo_green,
                   "Vehicle"            = waymo_blue)
waymo_anomaly <- c("No"                 = waymo_green, "Yes"                = waymo_navy)
waymo_feat    <- c("Speed"              = waymo_blue,  "Acceleration"       = waymo_green,
                   "YawRate"            = waymo_grey)

# --- Waymo ggplot Theme ---
waymo_theme <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
  theme(
    text             = element_text(family = "LM Roman", color = waymo_navy),
    plot.title       = element_text(face = "bold", size = base_size + 2, color = waymo_navy),
    plot.subtitle    = element_text(size = base_size - 1, color = waymo_grey),
    axis.text        = element_text(color = waymo_navy),
    axis.title       = element_text(color = waymo_navy),
    panel.grid.major = element_line(color = "#E8EAED"),
    panel.grid.minor = element_blank(),
    legend.text      = element_text(color = waymo_navy),
    legend.title     = element_text(color = waymo_navy, face = "bold"),
    plot.background  = element_rect(fill = waymo_white, color = NA),
    panel.background = element_rect(fill = waymo_white, color = NA),
    strip.text       = element_text(color = waymo_navy, face = "bold")
  )
}
```

```{r}
data <- read.csv("/Users/pavelstepanov/Library/CloudStorage/OneDrive-UniversityofMiami/Academic/Spring 2026/CSC 642/research/womd-unsupervised-anomaly-detection/data/waymo_dataset.csv")
cat("Total rows in raw dataset:", nrow(data), "\n")
```

```{r}
# Pre-processing
data_clean <- data %>%
  filter(Speed != 0) %>%
  filter(Valid_Current == "True") %>%
  drop_na(Speed, Acceleration, YawRate, Length, Width)

data_clean$AgentType <- as.factor(data_clean$AgentType)
cat("Final cleaned dataset size:", nrow(data_clean), "\n")
```

---

# 1. Correlation EDA

*Starting with EDA to understand feature relationships before modeling.*

```{r}
kinematic_vars <- c("Speed", "Acceleration", "YawRate", "Length", "Width")
cor_matrix <- cor(data_clean[, kinematic_vars], use = "complete.obs")
cat("=== Pearson Correlation Matrix ===\n")
print(round(cor_matrix, 3))
```

```{r}
# Correlation heatmap
cor_df <- as.data.frame(cor_matrix) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

dir.create("../../docs/images", recursive = TRUE, showWarnings = FALSE)

p_heatmap <- ggplot(cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = waymo_white, linewidth = 0.5) +
  geom_text(aes(label = round(Correlation, 2)), size = 3.5, fontface = "bold",
            color = waymo_navy, family = "LM Roman") +
  scale_fill_gradient2(low = waymo_blue, mid = waymo_white, high = waymo_green,
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Feature Correlation Heatmap — All Agent Types",
       subtitle = "YawRate is near-zero correlated with all other features",
       x = "", y = "") +
  waymo_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../docs/images/correlation_heatmap.png", plot = p_heatmap,
       width = 7, height = 6, dpi = 300, type = "cairo")
p_heatmap
```

```{r}
# Descriptive summary
summary(data_clean[, c("Speed", "Acceleration", "YawRate")])

p_speed <- ggplot(data_clean, aes(x = Speed, fill = AgentType)) +
  geom_histogram(bins = 40, color = waymo_white, linewidth = 0.2, alpha = 0.9) +
  scale_fill_manual(values = waymo_3col) +
  labs(title    = "Speed Distribution — All Agent Types",
       subtitle = "Moving agents only (Speed > 0); right-skewed across all classes",
       x        = "Speed (m/s)",
       y        = "Count",
       fill     = "Agent Type") +
  waymo_theme()

ggsave("../../docs/images/speed_distribution.png", plot = p_speed,
       width = 8, height = 5, dpi = 300, type = "cairo")
p_speed

cat("Speed skewness:", e1071::skewness(data_clean$Speed), "\n")
cat("99th percentile acceleration:", quantile(data_clean$Acceleration, 0.99), "\n")
```

---

# 2. Isolation Forest — Anomaly Detection

*YawRate's independence from other features makes it a strong candidate for anomaly isolation.*

```{r}
iso_model <- isolation.forest(
  data_clean[, c("Speed", "Acceleration", "YawRate")],
  ntrees   = 100,
  nthreads = 1
)

data_clean$AnomalyScore <- predict(
  iso_model,
  data_clean[, c("Speed", "Acceleration", "YawRate")]
)

score_threshold <- quantile(data_clean$AnomalyScore, 0.99)

data_clean$Is_Anomaly <- as.factor(
  ifelse(data_clean$AnomalyScore > score_threshold, "Yes", "No")
)

cat("Anomaly threshold (99th percentile):", score_threshold, "\n")
table(data_clean$Is_Anomaly)
```

```{r}
# Who are the anomalous agents?
anomalous_agents <- data_clean %>%
  filter(Is_Anomaly == "Yes") %>%
  select(ScenarioID, AgentID, AgentType, Speed, Acceleration,
         YawRate, Length, Width, AnomalyScore, IsInteractive) %>%
  arrange(desc(AnomalyScore))

cat("Total flagged:", nrow(anomalous_agents), "\n\n")

cat("=== Breakdown by Agent Type ===\n")
print(table(anomalous_agents$AgentType))

cat("\n=== Top 20 Most Anomalous Agents ===\n")
print(head(anomalous_agents, 20))

cat("\n=== Summary Statistics ===\n")
print(summary(anomalous_agents[, c("Speed", "Acceleration", "YawRate", "AnomalyScore")]))

cat("\n=== % flagged as Interactive by Waymo ===\n")
print(prop.table(table(anomalous_agents$IsInteractive)))
```

```{r}
# Fisher's Exact Test: is anomaly rate associated with AgentType?
contingency_table <- table(data_clean$AgentType, data_clean$Is_Anomaly)
fisher.test(contingency_table, simulate.p.value = TRUE)
```

```{r}
# Permutation feature importance
set.seed(42)
baseline_scores  <- data_clean$AnomalyScore
features_to_plot <- c("Speed", "Acceleration", "YawRate")

perm_importance <- sapply(features_to_plot, function(feat) {
  shuffled <- data_clean[, features_to_plot]
  shuffled[[feat]] <- sample(shuffled[[feat]])
  perm_scores <- predict(iso_model, shuffled)
  mean(abs(perm_scores - baseline_scores))
})

perm_df <- data.frame(
  Feature    = names(perm_importance),
  Importance = perm_importance
) %>% arrange(desc(Importance))

p_iso_importance <- ggplot(perm_df, aes(x = reorder(Feature, Importance),
                                        y = Importance, fill = Feature)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = waymo_feat) +
  labs(title    = "Isolation Forest: Permutation Feature Importance",
       subtitle = "Higher = shuffling this feature disrupts anomaly scores more",
       x = "", y = "Mean |Score Deviation|") +
  waymo_theme()

ggsave("../../docs/images/iso_forest_feature_importance.png", plot = p_iso_importance,
       width = 7, height = 5, dpi = 300, type = "cairo")
p_iso_importance

print(perm_df)
```

---

# 2b. Anomaly Detector Comparison — Isolation Forest vs. One-Class SVM vs. LOF

*The rubric calls for a head-to-head comparison; we evaluate three classical unsupervised detectors on the same kinematic feature set with the same 1% contamination rate, and benchmark each against Waymo's `IsInteractive` flag and the agent-type association test.*

```{r}
features    <- c("Speed", "Acceleration", "YawRate")
X           <- as.matrix(scale(data_clean[, features]))
n           <- nrow(X)
contam_frac <- 0.01
top_k       <- ceiling(contam_frac * n)

# --- Method 1: Isolation Forest (already fitted) ---
iso_pred_anomaly <- as.integer(data_clean$Is_Anomaly == "Yes")

# --- Method 2: One-Class SVM (RBF kernel, nu = 0.01) ---
set.seed(42)
ocsvm_model    <- svm(X, type = "one-classification",
                      kernel = "radial", nu = contam_frac, scale = FALSE)
ocsvm_decision <- predict(ocsvm_model, X, decision.values = TRUE)
ocsvm_scores   <- -as.numeric(attr(ocsvm_decision, "decision.values"))
ocsvm_thr      <- sort(ocsvm_scores, decreasing = TRUE)[top_k]
ocsvm_anomaly  <- as.integer(ocsvm_scores >= ocsvm_thr)

# --- Method 3: Local Outlier Factor (LOF, k = 20) ---
set.seed(42)
lof_scores  <- dbscan::lof(X, minPts = 20)
lof_thr     <- sort(lof_scores, decreasing = TRUE)[top_k]
lof_anomaly <- as.integer(lof_scores >= lof_thr)

# Bundle for downstream analysis
detector_flags <- list(
  "Isolation Forest" = iso_pred_anomaly,
  "One-Class SVM"    = ocsvm_anomaly,
  "LOF"              = lof_anomaly
)

cat("Number flagged per detector (target =", top_k, "):\n")
sapply(detector_flags, sum)
```

```{r}
# Pairwise overlap (Jaccard similarity of flagged sets)
jaccard <- function(a, b) {
  inter <- sum(a == 1 & b == 1)
  union <- sum(a == 1 | b == 1)
  if (union == 0) 0 else inter / union
}

methods <- names(detector_flags)
J <- matrix(NA, length(methods), length(methods),
            dimnames = list(methods, methods))
for (i in seq_along(methods))
  for (j in seq_along(methods))
    J[i, j] <- jaccard(detector_flags[[i]], detector_flags[[j]])

cat("=== Pairwise Jaccard Overlap of Flagged Anomalies ===\n")
print(round(J, 3))
```

```{r}
# Comparison metrics: agent-type association + IsInteractive precision
build_metrics <- function(flag_vec, name) {
  ct       <- table(data_clean$AgentType, factor(flag_vec, levels = c(0, 1)))
  fish_p   <- tryCatch(
    fisher.test(ct, simulate.p.value = TRUE, B = 2000)$p.value,
    error = function(e) NA_real_
  )
  flagged  <- which(flag_vec == 1)
  prec_int <- mean(data_clean$IsInteractive[flagged] == "True")
  pct_ped  <- mean(data_clean$AgentType[flagged] == "Pedestrian")
  pct_cyc  <- mean(data_clean$AgentType[flagged] == "Cyclist")
  data.frame(
    Method                  = name,
    N_Flagged               = sum(flag_vec),
    Fisher_p                = fish_p,
    Precision_IsInteractive = prec_int,
    Pct_Pedestrian          = pct_ped,
    Pct_Cyclist             = pct_cyc
  )
}

comparison_df <- bind_rows(lapply(methods, function(m)
  build_metrics(detector_flags[[m]], m)
))

cat("=== Detector Comparison Table ===\n")
print(comparison_df, digits = 4)

# Persist to CSV for the LaTeX paper
write.csv(comparison_df,
          "../../docs/submission/final-01/anomaly_comparison.csv",
          row.names = FALSE)
```

```{r}
# Visualize: bar chart of Fisher's test -log10(p) and Pedestrian % per detector
plot_df <- comparison_df %>%
  mutate(NegLogP = -log10(pmax(Fisher_p, 1e-4))) %>%
  select(Method, NegLogP, Pct_Pedestrian, Pct_Cyclist) %>%
  pivot_longer(-Method, names_to = "Metric", values_to = "Value")

p_compare <- ggplot(plot_df, aes(x = Method, y = Value, fill = Method)) +
  geom_col(show.legend = FALSE, color = waymo_white, linewidth = 0.3) +
  facet_wrap(~ Metric, scales = "free_y",
             labeller = labeller(Metric = c(
               NegLogP        = "-log10(Fisher p)",
               Pct_Pedestrian = "Fraction Pedestrian",
               Pct_Cyclist    = "Fraction Cyclist"
             ))) +
  scale_fill_manual(values = c("Isolation Forest" = waymo_blue,
                               "One-Class SVM"    = waymo_green,
                               "LOF"              = waymo_grey)) +
  labs(title    = "Unsupervised Anomaly Detector Comparison",
       subtitle = "Same 1% contamination on Speed/Acceleration/YawRate",
       x = "", y = "") +
  waymo_theme() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggsave("../../docs/submission/final-01/images/anomaly_method_comparison.png",
       plot = p_compare, width = 9, height = 4, dpi = 300, type = "cairo")
p_compare
```

```{r}
# Quick winner summary for the paper text
winner_idx <- which.min(comparison_df$Fisher_p)
cat("Best by Fisher's p:", comparison_df$Method[winner_idx],
    "(p =", comparison_df$Fisher_p[winner_idx], ")\n")

winner_idx2 <- which.max(comparison_df$Precision_IsInteractive)
cat("Best by IsInteractive precision:", comparison_df$Method[winner_idx2],
    "(prec =", round(comparison_df$Precision_IsInteractive[winner_idx2], 3), ")\n")
```

---

# 3. Linear Regression — Motion Prediction

*Does YawRate, identified as the key anomaly signal, also predict future displacement?*

```{r}
# Base cleaned dataset
base_data <- data %>%
  filter(!is.na(Speed), !is.na(Acceleration), !is.na(YawRate),
         !is.na(Vel_X), !is.na(Vel_Y), !is.na(FutureDist_5s))

data_with_stationary <- base_data
data_no_stationary   <- subset(base_data, Speed > 0)

cat("With stationary:", nrow(data_with_stationary), "\n")
cat("Without stationary:", nrow(data_no_stationary), "\n")

# Speed by AgentType (moving only)
cat("\nMean speed by agent type (moving agents):\n")
aggregate(Speed ~ AgentType, data = data_no_stationary, mean)
```

```{r}
run_model <- function(dataset) {
  set.seed(123)
  n           <- nrow(dataset)
  train_index <- sample(1:n, size = 0.8 * n)
  train       <- dataset[train_index, ]
  test        <- dataset[-train_index, ]
  control     <- trainControl(method = "cv", number = 5)

  model <- train(
    FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
    data      = train,
    method    = "lm",
    trControl = control
  )

  pred <- predict(model, newdata = test)
  rmse <- sqrt(mean((pred - test$FutureDist_5s)^2))
  r2   <- 1 - sum((pred - test$FutureDist_5s)^2) /
              sum((test$FutureDist_5s - mean(test$FutureDist_5s))^2)

  list(RMSE = rmse, R2 = r2, CV_Results = model$results)
}

results_with    <- run_model(data_with_stationary)
results_without <- run_model(data_no_stationary)

cat("=== With Stationary ===\n");    print(results_with)
cat("\n=== Without Stationary ===\n"); print(results_without)
```

## Regression Explainability

```{r}
lm_with    <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                 data = data_with_stationary)
lm_without <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                 data = data_no_stationary)

summary(lm_with)
summary(lm_without)
```

```{r}
# Coefficient plot with 95% CI — both datasets overlaid
extract_coefs <- function(model, label) {
  ci  <- confint(model)
  est <- coef(model)
  data.frame(Term = rownames(ci), Estimate = est,
             Lower = ci[,1], Upper = ci[,2], Dataset = label,
             stringsAsFactors = FALSE)
}

coef_df <- bind_rows(
  extract_coefs(lm_with,    "With Stationary"),
  extract_coefs(lm_without, "Without Stationary")
) %>% filter(Term != "(Intercept)")

p_coef_ci <- ggplot(coef_df, aes(x = reorder(Term, Estimate), y = Estimate,
                    color = Dataset, shape = Dataset)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper),
                position = position_dodge(width = 0.5), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = waymo_grey) +
  coord_flip() +
  scale_color_manual(values = waymo_2col) +
  labs(title    = "Regression Coefficients with 95% CI",
       subtitle = "With vs. without stationary agents",
       x = "", y = "Coefficient (metres per unit)") +
  waymo_theme()

ggsave("../../docs/images/regression_coefficients_ci.png", plot = p_coef_ci,
       width = 8, height = 5, dpi = 300, type = "cairo")
p_coef_ci
```
