---
title: "Waymo Project"
output: html_document
date: "2026-03-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Autonomous vehicles (AVs) must safely navigate dynamic urban environments involving vehicles, cyclists, and pedestrians. Two major challenges in this setting are detecting anomalous driving behavior and predicting short-term future motion of surrounding agents.

This project addresses both challenges using trajectory data derived from the Waymo Open Dataset. Specifically, we aim to:

- Detect anomalous motion patterns using unsupervised learning.
- Predict 5-second future displacement using supervised regression.
- Compare results when stationary agents are included versus removed.

---

# State-of-the-Art

Modern autonomous driving systems rely heavily on:

- Trajectory forecasting models (Kalman Filters, LSTMs, Transformers)
- Unsupervised anomaly detection (Isolation Forest, One-Class SVM)
- Multi-agent interaction models

Short-horizon forecasting (3–5 seconds) is often modeled using kinematic state variables. Isolation Forest remains widely used due to its scalability and minimal distributional assumptions.

---

# Materials and Methods

## Dataset

For each agent, kinematic features are extracted at the prediction horizon (t = 0). The feature set includes:

- **Speed (v)**: Magnitude of the velocity vector (m/s).
- **Acceleration (a)**: Rate of change of speed (m/s²).
- **Yaw Rate (ψ̇)**: Angular velocity representing the rate of change of heading (rad/s).

## Pipeline

1. Raw Waymo Motion Dataset extracted using Python (see `src/python/waymo.ipynb`).
2. Dataset exported as CSV.
3. Correlation EDA → Isolation Forest anomaly detection → Linear regression for motion prediction.

## Pre-processing

- Removed invalid frames (`Valid_Current != True`)
- Removed stationary agents (`Speed > 0`)
- Removed missing values in predictors

---

# Code Implementation

```{r}
library(tidyverse)
library(caret)
library(isotree)
library(e1071)
library(gridExtra)
library(car)
set.seed(42)
```

```{r}
data <- read.csv("/Users/justincabrera/Documents/Statistical Learning/Datasets/waymo_final_dataset.csv")
cat("Total rows in raw dataset:", nrow(data), "\n")
```

```{r}
# Pre-processing
data_clean <- data %>%
  filter(Speed != 0) %>%
  filter(Valid_Current == "True") %>%
  drop_na(Speed, Acceleration, YawRate, Length, Width)

data_clean$AgentType <- as.factor(data_clean$AgentType)
cat("Final cleaned dataset size:", nrow(data_clean), "\n")
```

---

# 1. Correlation EDA

*Starting with EDA to understand feature relationships before modeling.*

```{r}
kinematic_vars <- c("Speed", "Acceleration", "YawRate", "Length", "Width")
cor_matrix <- cor(data_clean[, kinematic_vars], use = "complete.obs")
cat("=== Pearson Correlation Matrix ===\n")
print(round(cor_matrix, 3))
```

```{r}
# Correlation heatmap
cor_df <- as.data.frame(cor_matrix) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

dir.create("../../docs/images", recursive = TRUE, showWarnings = FALSE)

p_heatmap <- ggplot(cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(Correlation, 2)), size = 3.5, fontface = "bold") +
  scale_fill_gradient2(low = "tomato", mid = "white", high = "steelblue",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Feature Correlation Heatmap — All Agent Types",
       subtitle = "YawRate is near-zero correlated with all other features",
       x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../docs/images/correlation_heatmap.png", plot = p_heatmap,
       width = 7, height = 6, dpi = 300)
p_heatmap
```

```{r}
# Descriptive summary
summary(data_clean[, c("Speed", "Acceleration", "YawRate")])

hist(data_clean$Speed,
     main = "Speed Distribution — All Agent Types",
     xlab = "Speed (m/s)",
     col  = "skyblue",
     breaks = 40)

cat("Speed skewness:", e1071::skewness(data_clean$Speed), "\n")
cat("99th percentile acceleration:", quantile(data_clean$Acceleration, 0.99), "\n")
```

---

# 2. Isolation Forest — Anomaly Detection

*YawRate's independence from other features makes it a strong candidate for anomaly isolation.*

```{r}
iso_model <- isolation.forest(
  data_clean[, c("Speed", "Acceleration", "YawRate")],
  ntrees   = 100,
  nthreads = 1
)

data_clean$AnomalyScore <- predict(
  iso_model,
  data_clean[, c("Speed", "Acceleration", "YawRate")]
)

score_threshold <- quantile(data_clean$AnomalyScore, 0.99)

data_clean$Is_Anomaly <- as.factor(
  ifelse(data_clean$AnomalyScore > score_threshold, "Yes", "No")
)

cat("Anomaly threshold (99th percentile):", score_threshold, "\n")
table(data_clean$Is_Anomaly)
```

```{r}
# Who are the anomalous agents?
anomalous_agents <- data_clean %>%
  filter(Is_Anomaly == "Yes") %>%
  select(ScenarioID, AgentID, AgentType, Speed, Acceleration,
         YawRate, Length, Width, AnomalyScore, IsInteractive) %>%
  arrange(desc(AnomalyScore))

cat("Total flagged:", nrow(anomalous_agents), "\n\n")

cat("=== Breakdown by Agent Type ===\n")
print(table(anomalous_agents$AgentType))

cat("\n=== Top 20 Most Anomalous Agents ===\n")
print(head(anomalous_agents, 20))

cat("\n=== Summary Statistics ===\n")
print(summary(anomalous_agents[, c("Speed", "Acceleration", "YawRate", "AnomalyScore")]))

cat("\n=== % flagged as Interactive by Waymo ===\n")
print(prop.table(table(anomalous_agents$IsInteractive)))
```

```{r}
# Fisher's Exact Test: is anomaly rate associated with AgentType?
contingency_table <- table(data_clean$AgentType, data_clean$Is_Anomaly)
fisher.test(contingency_table, simulate.p.value = TRUE)
```

## Isolation Forest Explainability

```{r}
# Feature distributions: normal vs anomalous
data_clean %>%
  select(Speed, Acceleration, YawRate, Is_Anomaly) %>%
  pivot_longer(-Is_Anomaly, names_to = "Feature", values_to = "Value") %>%
  ggplot(aes(x = Is_Anomaly, y = Value, fill = Is_Anomaly)) +
  geom_violin(alpha = 0.6, trim = TRUE) +
  geom_boxplot(width = 0.12, outlier.size = 0.8, alpha = 0.8) +
  facet_wrap(~Feature, scales = "free_y") +
  scale_fill_manual(values = c("No" = "#4CAF50", "Yes" = "#F44336")) +
  labs(title = "Feature Distributions: Normal vs. Anomalous Agents",
       x = "Is Anomaly", y = "Value", fill = "Anomaly") +
  theme_minimal()
```

```{r}
# Anomaly score vs each feature with LOESS
features_to_plot <- c("Speed", "Acceleration", "YawRate")

plots <- lapply(features_to_plot, function(feat) {
  ggplot(data_clean, aes_string(x = feat, y = "AnomalyScore", color = "Is_Anomaly")) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(aes(group = 1), method = "loess", se = FALSE,
                color = "black", linewidth = 0.8) +
    geom_hline(yintercept = score_threshold, linetype = "dashed", color = "red") +
    scale_color_manual(values = c("No" = "#4CAF50", "Yes" = "#F44336")) +
    labs(title = paste("Anomaly Score vs.", feat),
         subtitle = "Red dashed = 99th percentile threshold") +
    theme_minimal() +
    theme(legend.position = "none")
})

grid.arrange(grobs = plots, ncol = 2,
             top = "Isolation Forest: Score Sensitivity per Feature")
```

```{r}
# Permutation feature importance
set.seed(42)
baseline_scores <- data_clean$AnomalyScore

perm_importance <- sapply(features_to_plot, function(feat) {
  shuffled <- data_clean[, features_to_plot]
  shuffled[[feat]] <- sample(shuffled[[feat]])
  perm_scores <- predict(iso_model, shuffled)
  mean(abs(perm_scores - baseline_scores))
})

perm_df <- data.frame(
  Feature    = names(perm_importance),
  Importance = perm_importance
) %>% arrange(desc(Importance))

p_iso_importance <- ggplot(perm_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title    = "Isolation Forest: Permutation Feature Importance",
       subtitle = "Higher = shuffling this feature disrupts anomaly scores more",
       x = "", y = "Mean |Score Deviation|") +
  theme_minimal()

ggsave("../../docs/images/iso_forest_feature_importance.png", plot = p_iso_importance,
       width = 7, height = 5, dpi = 300)
p_iso_importance

print(perm_df)
```

```{r}
# Anomaly rate and mean score by AgentType
agent_summary <- data_clean %>%
  group_by(AgentType) %>%
  summarise(Count        = n(),
            Anomaly_Rate = mean(Is_Anomaly == "Yes"),
            Mean_Score   = mean(AnomalyScore),
            .groups      = "drop")

p1 <- ggplot(agent_summary, aes(x = AgentType, y = Anomaly_Rate, fill = AgentType)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Anomaly Rate by Agent Type", x = "", y = "% Flagged") +
  theme_minimal()

p2 <- ggplot(agent_summary, aes(x = AgentType, y = Mean_Score, fill = AgentType)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Mean Anomaly Score by Agent Type", x = "", y = "Mean Score") +
  theme_minimal()

g_agent <- arrangeGrob(p1, p2, ncol = 2)
ggsave("../../docs/images/anomaly_rate_by_agent_type.png", plot = g_agent,
       width = 10, height = 5, dpi = 300)
grid.arrange(p1, p2, ncol = 2)
print(agent_summary)
```

---

# 3. Linear Regression — Motion Prediction

*Does YawRate, identified as the key anomaly signal, also predict future displacement?*

```{r}
# Base cleaned dataset
base_data <- data %>%
  filter(!is.na(Speed), !is.na(Acceleration), !is.na(YawRate),
         !is.na(Vel_X), !is.na(Vel_Y), !is.na(FutureDist_5s))

data_with_stationary <- base_data
data_no_stationary   <- subset(base_data, Speed > 0)

cat("With stationary:", nrow(data_with_stationary), "\n")
cat("Without stationary:", nrow(data_no_stationary), "\n")

# Speed by AgentType (moving only)
cat("\nMean speed by agent type (moving agents):\n")
aggregate(Speed ~ AgentType, data = data_no_stationary, mean)
```

```{r}
run_model <- function(dataset) {
  set.seed(123)
  n           <- nrow(dataset)
  train_index <- sample(1:n, size = 0.8 * n)
  train       <- dataset[train_index, ]
  test        <- dataset[-train_index, ]
  control     <- trainControl(method = "cv", number = 5)

  model <- train(
    FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
    data      = train,
    method    = "lm",
    trControl = control
  )

  pred <- predict(model, newdata = test)
  rmse <- sqrt(mean((pred - test$FutureDist_5s)^2))
  r2   <- 1 - sum((pred - test$FutureDist_5s)^2) /
              sum((test$FutureDist_5s - mean(test$FutureDist_5s))^2)

  list(RMSE = rmse, R2 = r2, CV_Results = model$results)
}

results_with    <- run_model(data_with_stationary)
results_without <- run_model(data_no_stationary)

cat("=== With Stationary ===\n");    print(results_with)
cat("\n=== Without Stationary ===\n"); print(results_without)
```

## Regression Explainability

```{r}
lm_with    <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                 data = data_with_stationary)
lm_without <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                 data = data_no_stationary)

summary(lm_with)
summary(lm_without)
```

```{r}
# Coefficient plot with 95% CI — both datasets overlaid
extract_coefs <- function(model, label) {
  ci  <- confint(model)
  est <- coef(model)
  data.frame(Term = rownames(ci), Estimate = est,
             Lower = ci[,1], Upper = ci[,2], Dataset = label,
             stringsAsFactors = FALSE)
}

coef_df <- bind_rows(
  extract_coefs(lm_with,    "With Stationary"),
  extract_coefs(lm_without, "Without Stationary")
) %>% filter(Term != "(Intercept)")

p_coef_ci <- ggplot(coef_df, aes(x = reorder(Term, Estimate), y = Estimate,
                    color = Dataset, shape = Dataset)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper),
                position = position_dodge(width = 0.5), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(title    = "Regression Coefficients with 95% CI",
       subtitle = "With vs. without stationary agents",
       x = "", y = "Coefficient (metres per unit)") +
  theme_minimal()

ggsave("../../docs/images/regression_coefficients_ci.png", plot = p_coef_ci,
       width = 8, height = 5, dpi = 300)
p_coef_ci
```

```{r}
# Standardized beta coefficients
scale_df <- function(df) {
  df %>% mutate(across(c(Speed, Acceleration, YawRate, Vel_X, Vel_Y, FutureDist_5s),
                       ~ as.numeric(scale(.))))
}

lm_with_std    <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                     data = scale_df(data_with_stationary))
lm_without_std <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                     data = scale_df(data_no_stationary))

std_coef_df <- bind_rows(
  data.frame(Term = names(coef(lm_with_std)[-1]),
             Beta = coef(lm_with_std)[-1], Dataset = "With Stationary"),
  data.frame(Term = names(coef(lm_without_std)[-1]),
             Beta = coef(lm_without_std)[-1], Dataset = "Without Stationary")
)

p_std_coef <- ggplot(std_coef_df, aes(x = reorder(Term, abs(Beta)), y = Beta, fill = Dataset)) +
  geom_col(position = "dodge") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title    = "Standardized Coefficients (β)",
       subtitle = "Magnitude = relative importance; sign = direction of effect",
       x = "", y = "β") +
  theme_minimal()

ggsave("../../docs/images/regression_feature_importance.png", plot = p_std_coef,
       width = 7, height = 5, dpi = 300)
p_std_coef

print(std_coef_df)
```


```{r}
# VIF — multicollinearity check
cat("=== VIF: With Stationary ===\n");    print(vif(lm_with))
cat("\n=== VIF: Without Stationary ===\n"); print(vif(lm_without))
```