---
title: "Waymo Project"
output: html_document
date: "2026-03-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Test
# Introduction

Autonomous vehicles (AVs) must safely navigate dynamic urban environments involving vehicles, cyclists, and pedestrians. Two major challenges in this setting are detecting anomalous driving behavior and predicting short-term future motion of surrounding agents. 

This project addresses both challenges using trajectory data derived from the Waymo Open Dataset. Specifically, we aim to:

- Detect anomalous motion patterns using unsupervised learning.
- Predict 5-second future displacement using supervised regression.
- Compare results when stationary agents are included versus removed.

Understanding both behavioral irregularities and short-term motion forecasting is critical in pedestrian-heavy, low-speed environments where proactive planning improves safety.

---

# State-of-the-Art

Modern autonomous driving systems rely heavily on:

- Trajectory forecasting models (Kalman Filters, LSTMs, Transformers)
- Unsupervised anomaly detection (Isolation Forest, One-Class SVM)
- Multi-agent interaction models

Short-horizon forecasting (3–5 seconds) is often modeled using kinematic state variables. Isolation Forest remains widely used due to its scalability and minimal distributional assumptions.

This project implements classical machine learning techniques to establish strong baseline performance before considering deep learning extensions.

---

# Materials and Methods

## Dataset

### A. Feature Extraction

For each agent, kinematic features are extracted at the prediction horizon (t = 0) to capture the instantaneous motion state during periods of peak interaction density. These features describe both linear and angular motion dynamics.

The feature set includes:

- **Speed (v)**: Magnitude of the velocity vector (m/s).
- **Acceleration (a)**: Rate of change of speed (m/s²).
- **Yaw Rate (ψ̇)**: Angular velocity representing the rate of change of heading (rad/s).
- **Lateral G-Force**: Computed as |v × ψ̇|, serving as a proxy for cornering intensity and lateral maneuver aggressiveness.

These variables provide a compact yet expressive representation of instantaneous driving behavior and are used for both anomaly detection and motion prediction tasks.

See python notebook steps 1-4 for preprocessing: https://github.com/pstepanovum/womd-unsupervised-anomaly-detection/blob/main/src/python/waymo.ipynb

---

### B. Tooling and Data Pipeline

All statistical analysis, visualization, and modeling are performed in **R**. Python is used exclusively as a preprocessing tool to extract and structure the raw Waymo Motion Dataset into a tabular format suitable for statistical analysis.

The processing pipeline follows these steps:

1. Raw Waymo Motion Dataset extracted using Python.
2. Relevant kinematic features structured into tabular format.
3. Dataset exported as a comma-separated values (CSV) file.
4. Statistical analysis, visualization, dimensionality reduction, and anomaly detection conducted entirely in R.

This separation ensures reproducibility while leveraging Python for data engineering and R for statistical modeling and analysis.

The dataset contains:

- AgentType (Vehicle, Cyclist, Pedestrian)
- Speed
- Acceleration
- YawRate
- Velocity components (Vel_X, Vel_Y)
- FutureDist_5s (5-second future displacement)

### Pre-processing

- Removed invalid frames (`Valid_Current != True`)
- Removed missing values in predictors and target
- Created two datasets:
  - With stationary agents
  - Without stationary agents (Speed > 0)
- Converted categorical variables to factors
- Split data 80% training / 20% test

---

## Methods

### 1. Anomaly Detection (Unsupervised)

Method:
Isolation Forest (100 trees)

Features:
- Speed
- Acceleration
- YawRate

Top 1% of anomaly scores were labeled as anomalous.

Association between AgentType and anomaly was tested using Fisher’s Exact Test.

---

### 2. Motion Prediction (Supervised Regression)

Target:
FutureDist_5s

Predictors:
- Speed
- Acceleration
- YawRate
- Vel_X
- Vel_Y

Method:
Linear regression via caret.

Two scenarios were evaluated:
1. With stationary agents
2. Without stationary agents

---

# Evaluation

Dataset was split into:

- 80% Training
- 20% Test (Hold-out)

5-fold cross-validation was performed on the training set.

Metrics:

Classification:
- Confusion Matrix
- ROC (AUC)
- Sensitivity / Specificity

Regression:
- RMSE
- R²

---

# Results

## Descriptive Analysis

When including stationary agents, vehicles showed artificially lower average speeds due to stopping behavior at intersections.

After removing stationary agents:

- Vehicles: ~6.74 m/s  
- Cyclists: ~3.92 m/s  
- Pedestrians: ~1.07 m/s  

These values align with real-world expectations.

---

## Anomaly Detection Results

Isolation Forest identified approximately 1% of observations as anomalous. These correspond to extreme acceleration, sharp yaw changes, or unusually high speeds.

This supports the use of anomaly detection as a safety monitoring tool in AV systems.

---

## Motion Prediction Results

Regression models demonstrated strong predictive performance for 5-second displacement.

Removing stationary agents improved interpretability and reduced bias from zero-speed clustering.

These findings suggest that short-term forecasting can be effectively achieved using current kinematic state variables.

---

# Discussion

This project demonstrates that:

- Kinematic features strongly determine short-term motion.
- Isolation Forest effectively detects rare behavioral patterns.
- Including stationary agents significantly alters descriptive statistics.
- Combining anomaly detection with motion prediction provides a robust safety modeling framework.

In pedestrian-heavy urban environments, anomaly detection flags irregular behavior while motion forecasting anticipates future positions. Together, they support proactive planning in autonomous vehicles.

---

# Code Implementation

```{r}
library(tidyverse)
library(caret)
library(isotree)
library(e1071)
library(randomForest)
library(pdp)
library(car)
library(gridExtra)
set.seed(42)
```

```{r}
    # Load Data
data<-read.csv("/Users/justincabrera/Documents/Statistical Learning/Datasets/waymo_final_dataset.csv")  
cat("Total rows in raw dataset:", nrow(data))
  
```
    
    
    
```{r}
# Pre-processing
data_filtered <- data %>%
  filter(Speed != 0)

data_clean <- data_filtered %>%
  filter(Valid_Current == "True") %>%
  drop_na(Speed, Acceleration, YawRate, Length, Width)

data_clean$AgentType <- as.factor(data_clean$AgentType)

cat("Final cleaned dataset size:", nrow(data_clean))
```

``` {r}
# Descriptive Analysis
summary(data_clean$Speed)

hist(data_clean$Speed,
     main = "Vehicle Speed Distribution",
     xlab = "Speed (m/s)",
     col = "skyblue")

speed_skewness <- skewness(data_clean$Speed)
cat("Speed Skewness:", speed_skewness)

accel_threshold <- quantile(data_clean$Acceleration, 0.99)
cat("99th Percentile Acceleration:", accel_threshold)
```



``` {r}
# Isolation Forest
iso_model <- isolation.forest(
  data_clean[, c("Speed", "Acceleration", "YawRate")],
  ntrees = 100,
  nthreads = 1
)

data_clean$AnomalyScore <- predict(
  iso_model,
  data_clean[, c("Speed", "Acceleration", "YawRate")]
)

score_threshold <- quantile(data_clean$AnomalyScore, 0.99)

data_clean$Is_Anomaly <- as.factor(
  ifelse(data_clean$AnomalyScore > score_threshold, "Yes", "No")
)

table(data_clean$Is_Anomaly)
```

``` {r}
# --- Who are the anomalous agents? ---
anomalous_agents <- data_clean %>%
  filter(Is_Anomaly == "Yes") %>%
  select(ScenarioID, AgentID, AgentType, Speed, Acceleration, YawRate, Length, Width, AnomalyScore, IsInteractive) %>%
  arrange(desc(AnomalyScore))

cat("Total anomalous agents flagged:", nrow(anomalous_agents), "\n\n")

cat("=== Breakdown by Agent Type ===\n")
print(table(anomalous_agents$AgentType))

cat("\n=== Top 20 Most Anomalous Agents ===\n")
print(head(anomalous_agents, 20))

cat("\n=== Summary Statistics of Anomalous Agents ===\n")
print(summary(anomalous_agents[, c("Speed", "Acceleration", "YawRate", "AnomalyScore")]))

cat("\n=== What % of anomalous agents are flagged as Interactive by Waymo? ===\n")
print(prop.table(table(anomalous_agents$IsInteractive)))
```

``` {r}
# Statistical Test
contingency_table <- table(data_clean$AgentType,
                           data_clean$Is_Anomaly)

fisher_results <- fisher.test(contingency_table,
                              simulate.p.value = TRUE)

fisher_results
```

---

## Isolation Forest Explainability

``` {r}
# --- Feature distributions: Anomalous vs. Normal ---
data_clean %>%
  select(Speed, Acceleration, YawRate, Is_Anomaly) %>%
  pivot_longer(-Is_Anomaly, names_to = "Feature", values_to = "Value") %>%
  ggplot(aes(x = Is_Anomaly, y = Value, fill = Is_Anomaly)) +
  geom_violin(alpha = 0.6, trim = TRUE) +
  geom_boxplot(width = 0.12, outlier.size = 0.8, alpha = 0.8) +
  facet_wrap(~Feature, scales = "free_y") +
  scale_fill_manual(values = c("No" = "#4CAF50", "Yes" = "#F44336")) +
  labs(
    title = "Kinematic Feature Distributions: Normal vs. Anomalous Agents",
    subtitle = "Anomalies are concentrated at the tails of each distribution",
    x = "Is Anomaly", y = "Value", fill = "Anomaly"
  ) +
  theme_minimal()
```

``` {r}
# --- Anomaly score vs. each feature (scatter with LOESS) ---
features_to_plot <- c("Speed", "Acceleration", "YawRate")

plots <- lapply(features_to_plot, function(feat) {
  ggplot(data_clean, aes_string(x = feat, y = "AnomalyScore",
                                color = "Is_Anomaly")) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(aes(group = 1), method = "loess", se = FALSE,
                color = "black", linewidth = 0.8) +
    geom_hline(yintercept = quantile(data_clean$AnomalyScore, 0.99),
               linetype = "dashed", color = "red") +
    scale_color_manual(values = c("No" = "#4CAF50", "Yes" = "#F44336")) +
    labs(title = paste("Anomaly Score vs.", feat),
         subtitle = "Red dashed line = 99th percentile threshold",
         x = feat, y = "Anomaly Score") +
    theme_minimal() +
    theme(legend.position = "none")
})

grid.arrange(grobs = plots, ncol = 2,
             top = "Isolation Forest: Score Sensitivity per Feature")
```

``` {r}
# --- Permutation feature importance for Isolation Forest ---
# Measures how much the anomaly score distribution shifts when a feature is shuffled
set.seed(42)
kinematic_features <- c("Speed", "Acceleration", "YawRate")
baseline_scores    <- data_clean$AnomalyScore

perm_importance <- sapply(kinematic_features, function(feat) {
  shuffled       <- data_clean[, kinematic_features]
  shuffled[[feat]] <- sample(shuffled[[feat]])
  perm_scores    <- predict(iso_model, shuffled)
  # Importance = mean absolute deviation from baseline when feature is destroyed
  mean(abs(perm_scores - baseline_scores))
})

perm_df <- data.frame(
  Feature    = names(perm_importance),
  Importance = perm_importance
) %>% arrange(desc(Importance))

ggplot(perm_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title    = "Isolation Forest: Permutation Feature Importance",
    subtitle = "Higher value = shuffling this feature disrupts anomaly scores more",
    x = "", y = "Mean |Score Deviation|"
  ) +
  theme_minimal()

print(perm_df)
```

``` {r}
# --- Anomaly rate and mean score by AgentType ---
agent_summary <- data_clean %>%
  group_by(AgentType) %>%
  summarise(
    Count        = n(),
    Anomaly_Rate = mean(Is_Anomaly == "Yes"),
    Mean_Score   = mean(AnomalyScore),
    .groups      = "drop"
  )

p1 <- ggplot(agent_summary, aes(x = AgentType, y = Anomaly_Rate, fill = AgentType)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Anomaly Rate by Agent Type", x = "", y = "% Flagged as Anomalous") +
  theme_minimal()

p2 <- ggplot(agent_summary, aes(x = AgentType, y = Mean_Score, fill = AgentType)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Mean Anomaly Score by Agent Type", x = "", y = "Mean Isolation Forest Score") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
print(agent_summary)
```

``` {r}
# Train/Test Split
trainIndex <- createDataPartition(data_clean$Is_Anomaly,
                                  p = .8,
                                  list = FALSE)

train_data <- data_clean[trainIndex, ]
test_data  <- data_clean[-trainIndex, ]
```

``` {r}
# Handle Class Imbalance
train_balanced <- downSample(
  x = train_data[, c("Length", "Width", "YawRate")],
  y = train_data$Is_Anomaly
)

names(train_balanced)[names(train_balanced) == "Class"] <- "Is_Anomaly"

table(train_balanced$Is_Anomaly)
```

``` {r}
# Random Forest Model
fit_control <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

model_rf <- train(
  Is_Anomaly ~ Length + Width + YawRate,
  data = train_balanced,
  method = "rf",
  metric = "ROC",
  trControl = fit_control
)

model_rf
```

``` {r}
# Evaluation
predictions <- predict(model_rf, test_data)

conf_matrix <- confusionMatrix(predictions,
                               test_data$Is_Anomaly)

conf_matrix
```

---

## Random Forest Explainability

``` {r}
# --- Variable Importance ---
rf_importance <- varImp(model_rf, scale = TRUE)

plot(rf_importance,
     main = "Random Forest: Variable Importance (Mean Decrease in Gini)")

# Print numeric values
print(rf_importance)
```

``` {r}
# --- Partial Dependence Plots (marginal effect of each feature on P(Anomaly)) ---
pd_yawrate <- partial(model_rf, pred.var = "YawRate",
                      which.class = "Yes", prob = TRUE, train = train_balanced)
pd_length  <- partial(model_rf, pred.var = "Length",
                      which.class = "Yes", prob = TRUE, train = train_balanced)
pd_width   <- partial(model_rf, pred.var = "Width",
                      which.class = "Yes", prob = TRUE, train = train_balanced)

p_yr <- autoplot(pd_yawrate) +
  labs(title = "PDP: Yaw Rate", subtitle = "Effect on P(Anomaly = Yes)",
       x = "YawRate (rad/s)", y = "Predicted P(Anomaly)") +
  theme_minimal()

p_ln <- autoplot(pd_length) +
  labs(title = "PDP: Agent Length", subtitle = "Effect on P(Anomaly = Yes)",
       x = "Length (m)", y = "Predicted P(Anomaly)") +
  theme_minimal()

p_wd <- autoplot(pd_width) +
  labs(title = "PDP: Agent Width", subtitle = "Effect on P(Anomaly = Yes)",
       x = "Width (m)", y = "Predicted P(Anomaly)") +
  theme_minimal()

grid.arrange(p_yr, p_ln, p_wd, ncol = 2,
             top = "Partial Dependence Plots: Marginal Feature Effects on Anomaly Probability")
```

``` {r}
# --- Two-way PDP: interaction between YawRate and Width ---
pd_2way <- partial(model_rf, pred.var = c("YawRate", "Width"),
                   which.class = "Yes", prob = TRUE, train = train_balanced)

plotPartial(pd_2way,
            main  = "2-Way PDP: YawRate × Width Interaction on P(Anomaly)",
            xlab  = "YawRate (rad/s)", ylab = "Width (m)",
            col.regions = heat.colors(100))
```

#Linear Regression on Future Distance 5s: 

```{r}
model_data <- data[, c(
  "AgentType",
  "Speed",
  "Acceleration",
  "YawRate",
  "Vel_X",
  "Vel_Y",
  "FutureDist_5s"
)]

# Remove rows missing FutureDist_5s
model_data <- subset(model_data, !is.na(FutureDist_5s))

```





```{r}
summary(model_data$Speed)
summary(model_data$FutureDist_5s)

aggregate(Speed ~ AgentType, data = model_data, mean)

```

```{r}

moving_data <- subset(data, Speed > 0)

aggregate(Speed ~ AgentType, data = moving_data, mean)
```

When including stationary agents, vehicles exhibited a lower mean speed than cyclists due to a high proportion of zero-velocity observations (e.g., vehicles stopped at intersections). After filtering for moving agents (Speed > 0), vehicles exhibited the highest average speed (6.74 m/s), followed by cyclists (3.92 m/s) and pedestrians (1.07 m/s), consistent with real-world expectations


```{r}


FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y

```

```{r}
# Base cleaned dataset (no NA in predictors or target)
base_data <- subset(data, 
                    !is.na(Speed) &
                    !is.na(Acceleration) &
                    !is.na(YawRate) &
                    !is.na(Vel_X) &
                    !is.na(Vel_Y) &
                    !is.na(FutureDist_5s))

# Dataset A: WITH stationary vehicles
data_with_stationary <- base_data

# Dataset B: WITHOUT stationary vehicles
data_no_stationary <- subset(base_data, Speed > 0)

dim(data_with_stationary)
dim(data_no_stationary)

```

```{r}

library(caret)

run_model <- function(dataset) {
  
  set.seed(123)
  
  n <- nrow(dataset)
  train_index <- sample(1:n, size = 0.8*n)
  
  train <- dataset[train_index, ]
  test  <- dataset[-train_index, ]
  
  control <- trainControl(method = "cv", number = 5)
  
  model <- train(
    FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
    data = train,
    method = "lm",
    trControl = control
  )
  
  # Test predictions
  pred <- predict(model, newdata = test)
  
  # RMSE
  rmse <- sqrt(mean((pred - test$FutureDist_5s)^2))
  
  # R-squared
  rss <- sum((pred - test$FutureDist_5s)^2)
  tss <- sum((test$FutureDist_5s - mean(test$FutureDist_5s))^2)
  r2 <- 1 - rss/tss
  
  return(list(
    RMSE = rmse,
    R2 = r2,
    CV_Results = model$results
  ))
}


```


```{r}

results_with <- run_model(data_with_stationary)
results_without <- run_model(data_no_stationary)

results_with
results_without
```

---

## Linear Regression Explainability

``` {r}
# Fit plain lm() models for explainability (caret wraps lm, so we refit for diagnostics)
lm_with    <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                 data = data_with_stationary)
lm_without <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                 data = data_no_stationary)

summary(lm_with)
summary(lm_without)
```

``` {r}
# --- Coefficient plot with 95% CI (both datasets overlaid) ---
extract_coefs <- function(model, label) {
  ci  <- confint(model)
  est <- coef(model)
  data.frame(
    Term     = rownames(ci),
    Estimate = est,
    Lower    = ci[, 1],
    Upper    = ci[, 2],
    Dataset  = label,
    stringsAsFactors = FALSE
  )
}

coef_df <- bind_rows(
  extract_coefs(lm_with,    "With Stationary"),
  extract_coefs(lm_without, "Without Stationary")
) %>% filter(Term != "(Intercept)")

ggplot(coef_df, aes(x = reorder(Term, Estimate), y = Estimate,
                    color = Dataset, shape = Dataset)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper),
                position = position_dodge(width = 0.5), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(
    title    = "Regression Coefficients with 95% Confidence Intervals",
    subtitle = "Comparing models with vs. without stationary agents",
    x = "", y = "Coefficient (metres per unit)"
  ) +
  theme_minimal()
```

``` {r}
# --- Standardized (beta) coefficients for relative importance ---
# Scale all continuous predictors to mean=0, SD=1 before fitting
std_formula <- FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y

scale_df <- function(df) {
  df %>% mutate(across(c(Speed, Acceleration, YawRate, Vel_X, Vel_Y, FutureDist_5s),
                       ~ as.numeric(scale(.))))
}

lm_with_std    <- lm(std_formula, data = scale_df(data_with_stationary))
lm_without_std <- lm(std_formula, data = scale_df(data_no_stationary))

std_coef_df <- bind_rows(
  data.frame(Term = names(coef(lm_with_std)[-1]),
             Beta = coef(lm_with_std)[-1],
             Dataset = "With Stationary"),
  data.frame(Term = names(coef(lm_without_std)[-1]),
             Beta = coef(lm_without_std)[-1],
             Dataset = "Without Stationary")
)

ggplot(std_coef_df, aes(x = reorder(Term, abs(Beta)), y = Beta, fill = Dataset)) +
  geom_col(position = "dodge") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title    = "Standardized Regression Coefficients (β)",
    subtitle = "Magnitude reflects relative importance; sign shows direction",
    x = "", y = "Standardized Coefficient (β)"
  ) +
  theme_minimal()

print(std_coef_df)
```

``` {r}
# --- Variance Inflation Factors (multicollinearity check) ---
cat("=== VIF: With Stationary ===\n")
print(vif(lm_with))

cat("\n=== VIF: Without Stationary ===\n")
print(vif(lm_without))
```

``` {r}
# --- Added-Variable (Partial Regression) Plots ---
# Shows the unique contribution of each predictor after accounting for all others.
# Points off the horizontal = leverage observations; slope = partial coefficient.
par(mfrow = c(1, 1))
avPlots(lm_without,
        main = "Added-Variable Plots (Moving Agents Only)",
        ask  = FALSE)
```

``` {r}
# --- Component + Residual Plots (non-linearity check) ---
# A non-linear smooth (pink) diverging from the straight line suggests a
# predictor needs a transformation (e.g., log-speed or speed²).
crPlots(lm_without,
        main = "Component + Residual Plots (Moving Agents Only)",
        ask  = FALSE)
```

