---
title: "Waymo — Vehicles Only: Isolation Forest + Regression"
output: html_document
date: "2026-04-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Repeats the full analysis pipeline restricted to **Vehicles only**. Pedestrians
and cyclists are excluded — their kinematic profiles (near-zero speed, high yaw
rate noise from sensor clipping) dominate anomaly scores and obscure
vehicle-specific dangerous behavior.

Flow: Correlation EDA → Isolation Forest → Linear Regression

---

# Setup

```{r}
library(tidyverse)
library(caret)
library(isotree)
library(e1071)
library(gridExtra)
library(car)
set.seed(42)
```

```{r}
data <- read.csv("/Users/justincabrera/Documents/Statistical Learning/Datasets/waymo_final_dataset.csv")
cat("Total rows in raw dataset:", nrow(data), "\n")
```

---

# Preprocessing — Vehicles Only

```{r}
vehicles <- data %>%
  filter(AgentType == "Vehicle") %>%
  filter(Valid_Current == "True") %>%
  filter(Speed > 0) %>%
  drop_na(Speed, Acceleration, YawRate, Length, Width)

cat("Vehicle agents after filtering:", nrow(vehicles), "\n")
summary(vehicles[, c("Speed", "Acceleration", "YawRate")])
```

```{r}
hist(vehicles$Speed,
     main = "Vehicle Speed Distribution (Moving Only)",
     xlab = "Speed (m/s)",
     col  = "steelblue",
     breaks = 40)

cat("99th percentile speed:", quantile(vehicles$Speed, 0.99), "m/s\n")
cat("99th percentile acceleration:", quantile(vehicles$Acceleration, 0.99), "m/s²\n")
cat("99th percentile |YawRate|:", quantile(abs(vehicles$YawRate), 0.99), "rad/s\n")
```

---

# 1. Correlation EDA — Vehicles Only

```{r}
kinematic_vars <- c("Speed", "Acceleration", "YawRate", "Length", "Width")
cor_matrix_v   <- cor(vehicles[, kinematic_vars], use = "complete.obs")
cat("=== Pearson Correlation Matrix — Vehicles ===\n")
print(round(cor_matrix_v, 3))
```

```{r}
# Correlation heatmap
cor_df_v <- as.data.frame(cor_matrix_v) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

ggplot(cor_df_v, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(Correlation, 2)), size = 3.5, fontface = "bold") +
  scale_fill_gradient2(low = "tomato", mid = "white", high = "steelblue",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title    = "Feature Correlation Heatmap — Vehicles Only",
       subtitle = "YawRate remains orthogonal to all other features",
       x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Pairs scatter plot
pairs(vehicles[, kinematic_vars],
      col  = adjustcolor("steelblue", alpha.f = 0.2),
      pch  = 16,
      cex  = 0.3,
      main = "Feature Pairs Plot — Vehicles Only")
```

```{r}
# Speed vs Acceleration colored by |YawRate|
vehicles %>%
  mutate(YawRate_Magnitude = abs(YawRate)) %>%
  ggplot(aes(x = Speed, y = Acceleration, color = YawRate_Magnitude)) +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_color_gradient(low = "steelblue", high = "red", name = "|YawRate|\n(rad/s)") +
  labs(title    = "Speed vs. Acceleration — colored by |YawRate|",
       subtitle = "Red = high angular velocity while accelerating/decelerating",
       x = "Speed (m/s)", y = "Acceleration (m/s²)") +
  theme_minimal()
```

---

# 2. Isolation Forest — Vehicles Only

```{r}
iso_vehicle <- isolation.forest(
  vehicles[, c("Speed", "Acceleration", "YawRate")],
  ntrees   = 100,
  nthreads = 1
)

vehicles$AnomalyScore <- predict(
  iso_vehicle,
  vehicles[, c("Speed", "Acceleration", "YawRate")]
)

score_threshold <- quantile(vehicles$AnomalyScore, 0.99)

vehicles$Is_Anomaly <- as.factor(
  ifelse(vehicles$AnomalyScore > score_threshold, "Yes", "No")
)

cat("Anomaly threshold (99th percentile):", score_threshold, "\n")
table(vehicles$Is_Anomaly)
```

```{r}
# Who are the anomalous vehicles?
anomalous_vehicles <- vehicles %>%
  filter(Is_Anomaly == "Yes") %>%
  select(ScenarioID, AgentID, AgentType, Speed, Acceleration,
         YawRate, Length, Width, AnomalyScore, IsInteractive) %>%
  arrange(desc(AnomalyScore))

cat("Total anomalous vehicles flagged:", nrow(anomalous_vehicles), "\n\n")
cat("=== Top 20 Most Anomalous Vehicles ===\n")
print(head(anomalous_vehicles, 20))

cat("\n=== % flagged as Interactive by Waymo ===\n")
print(prop.table(table(anomalous_vehicles$IsInteractive)))
```

## Isolation Forest Explainability

```{r}
# Feature distributions: normal vs anomalous vehicles
vehicles %>%
  select(Speed, Acceleration, YawRate, Is_Anomaly) %>%
  pivot_longer(-Is_Anomaly, names_to = "Feature", values_to = "Value") %>%
  ggplot(aes(x = Is_Anomaly, y = Value, fill = Is_Anomaly)) +
  geom_violin(alpha = 0.6, trim = TRUE) +
  geom_boxplot(width = 0.12, outlier.size = 0.8, alpha = 0.8) +
  facet_wrap(~Feature, scales = "free_y") +
  scale_fill_manual(values = c("No" = "#2196F3", "Yes" = "#F44336")) +
  labs(title    = "Vehicle Feature Distributions: Normal vs. Anomalous",
       subtitle = "Vehicles only — pedestrian noise removed",
       x = "Is Anomaly", y = "Value", fill = "Anomaly") +
  theme_minimal()
```

```{r}
# Anomaly score vs each feature with LOESS
features_to_plot <- c("Speed", "Acceleration", "YawRate")

plots <- lapply(features_to_plot, function(feat) {
  ggplot(vehicles, aes_string(x = feat, y = "AnomalyScore", color = "Is_Anomaly")) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(aes(group = 1), method = "loess", se = FALSE,
                color = "black", linewidth = 0.8) +
    geom_hline(yintercept = score_threshold, linetype = "dashed", color = "red") +
    scale_color_manual(values = c("No" = "#2196F3", "Yes" = "#F44336")) +
    labs(title = paste("Anomaly Score vs.", feat)) +
    theme_minimal() +
    theme(legend.position = "none")
})

grid.arrange(grobs = plots, ncol = 2,
             top = "Isolation Forest: Score vs. Feature (Vehicles Only)")
```

```{r}
# Permutation feature importance
set.seed(42)
baseline_scores <- vehicles$AnomalyScore

perm_importance <- sapply(features_to_plot, function(feat) {
  shuffled <- vehicles[, features_to_plot]
  shuffled[[feat]] <- sample(shuffled[[feat]])
  perm_scores <- predict(iso_vehicle, shuffled)
  mean(abs(perm_scores - baseline_scores))
})

perm_df <- data.frame(
  Feature    = names(perm_importance),
  Importance = perm_importance
) %>% arrange(desc(Importance))

ggplot(perm_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title    = "Isolation Forest: Permutation Feature Importance (Vehicles Only)",
       subtitle = "Which feature most drives anomaly detection for vehicles?",
       x = "", y = "Mean |Score Deviation|") +
  theme_minimal()

print(perm_df)
```

---

# 3. Linear Regression — Vehicle Motion Prediction

*Does YawRate, the dominant anomaly signal, also predict future displacement?*

```{r}
vehicle_reg <- data %>%
  filter(AgentType == "Vehicle") %>%
  filter(Speed > 0) %>%
  filter(Valid_Current == "True") %>%
  drop_na(Speed, Acceleration, YawRate, Vel_X, Vel_Y, FutureDist_5s)

cat("Vehicle regression dataset size:", nrow(vehicle_reg), "\n")
summary(vehicle_reg$FutureDist_5s)
```

```{r}
set.seed(123)
n           <- nrow(vehicle_reg)
train_idx   <- sample(1:n, size = 0.8 * n)
train_reg   <- vehicle_reg[train_idx, ]
test_reg    <- vehicle_reg[-train_idx, ]
control     <- trainControl(method = "cv", number = 5)

model_lm_v <- train(
  FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
  data      = train_reg,
  method    = "lm",
  trControl = control
)

pred_v <- predict(model_lm_v, newdata = test_reg)
rmse_v <- sqrt(mean((pred_v - test_reg$FutureDist_5s)^2))
r2_v   <- 1 - sum((pred_v - test_reg$FutureDist_5s)^2) /
              sum((test_reg$FutureDist_5s - mean(test_reg$FutureDist_5s))^2)

cat("RMSE:", rmse_v, "\n")
cat("R²:", r2_v, "\n")
print(model_lm_v$results)
```

## Regression Explainability

```{r}
lm_vehicle <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
                 data = vehicle_reg)
summary(lm_vehicle)
```

```{r}
# Coefficient plot with 95% CI
ci      <- confint(lm_vehicle)
coef_df <- data.frame(
  Term     = rownames(ci),
  Estimate = coef(lm_vehicle),
  Lower    = ci[, 1],
  Upper    = ci[, 2]
) %>% filter(Term != "(Intercept)")

ggplot(coef_df, aes(x = reorder(Term, Estimate), y = Estimate)) +
  geom_point(color = "steelblue", size = 3) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(title = "Regression Coefficients with 95% CI — Vehicles Only",
       x = "", y = "Coefficient (metres per unit)") +
  theme_minimal()
```

```{r}
# Standardized beta coefficients
scale_df <- function(df) {
  df %>% mutate(across(c(Speed, Acceleration, YawRate, Vel_X, Vel_Y, FutureDist_5s),
                       ~ as.numeric(scale(.))))
}

lm_std <- lm(FutureDist_5s ~ Speed + Acceleration + YawRate + Vel_X + Vel_Y,
             data = scale_df(vehicle_reg))

std_df <- data.frame(
  Term = names(coef(lm_std)[-1]),
  Beta = coef(lm_std)[-1]
) %>% arrange(desc(abs(Beta)))

ggplot(std_df, aes(x = reorder(Term, abs(Beta)), y = Beta, fill = Beta > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  labs(title    = "Standardized Coefficients (β) — Vehicles Only",
       subtitle = "Blue = positive effect, Red = negative effect on 5s displacement",
       x = "", y = "β") +
  theme_minimal()

print(std_df)
```

```{r}
# VIF — multicollinearity check
cat("=== VIF: Vehicle Regression ===\n")
print(vif(lm_vehicle))
```

```{r}
# Added-variable plots
avPlots(lm_vehicle, main = "Added-Variable Plots — Vehicles Only", ask = FALSE)
```

```{r}
# Component + residual plots
crPlots(lm_vehicle, main = "Component + Residual Plots — Vehicles Only", ask = FALSE)
```
